Probabilistic inference in dynamical systems

### 2024-04-26 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): add inference plots to concepts Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

>>>>

Now we can set up some sample data and perform inference on the latent variables

```{python}
# | label: fig-generate-test-data-for-deterministic-model-inference
# | code-fold: true
# | fig-cap: Test data

import arviz as az
from pyrovelocity.models._deterministic_inference import (
    generate_posterior_inference_data,
    generate_prior_inference_data,
    generate_test_data_for_deterministic_model_inference,
    plot_sample_phase_portraits,
    plot_sample_trajectories,
    plot_sample_trajectories_with_percentiles,
    print_inference_data_structure,
)

(
    times,
    data,
    num_genes,
    num_cells,
    num_timepoints,
    num_modalities,
) = generate_test_data_for_deterministic_model_inference(
    num_genes=1,
    num_cells=20,
    num_timepoints=1,
    num_modalities=2,
    noise_levels=(0.001, 0.001),
)

num_chains = 1
num_samples = 10
num_warmup = 10

idata_prior = generate_prior_inference_data(
    times=times,
    data=data,
    num_chains=num_chains,
    num_samples=num_samples,
    num_genes=num_genes,
    num_cells=num_cells,
    num_timepoints=num_timepoints,
    num_modalities=num_modalities,
)

idata_posterior = generate_posterior_inference_data(
    times=times,
    data=data,
    num_chains=num_chains,
    num_samples=num_samples,
    num_genes=num_genes,
    num_cells=num_cells,
    num_timepoints=num_timepoints,
    num_modalities=num_modalities,
    num_warmup=num_warmup,
)
```

We can plot the prior

```{python}
# | label: fig-transcription-splicing-degradation-prior
# | code-fold: true
# | fig-cap: Prior distributions for the transcription-splicing-degradation model.
light_gray = "#bcbcbc"
variables = ["initial_conditions", "gamma", "sigma"]
for var in variables:
    az.plot_posterior(
        idata_prior,
        var_names=[var],
        group="prior",
        kind="hist",
        color=light_gray,
        round_to=2,
    )
```

and posterior distributions for the latent variables

```{python}
# | label: fig-transcription-splicing-degradation-posterior
# | code-fold: true
# | fig-cap: Posterior distributions for the transcription-splicing-degradation model.
light_gray = "#bcbcbc"
variables = ["initial_conditions", "gamma", "sigma"]
for var in variables:
    az.plot_posterior(
        idata_posterior,
        var_names=[var],
        group="posterior",
        kind="hist",
        color=light_gray,
        round_to=2,
    )
```

We can plot posterior predictive trajectories

```{python}
# | label: fig-transcription-splicing-degradation-posterior-predictive-trajectories
# | code-fold: true
# | fig-cap: Posterior predictive trajectories for the transcription-splicing-degradation model.
figs = plot_sample_trajectories(
    idata=idata_posterior,
)
for fig in figs:
    fig.show()
```

trajectories in the phase portrait space

```{python}
# | label: fig-transcription-splicing-degradation-posterior-predictive-phase-portraits
# | code-fold: true
# | fig-cap: Posterior predictive phase portraits for the transcription-splicing-degradation model.
figs = plot_sample_phase_portraits(
    idata=idata_posterior,
    colormap_name="RdBu",
)
for fig in figs:
    fig.show()
```

as well as estimates for the distribution of the pre-mRNA and mRNA

```{python}
# | label: fig-transcription-splicing-degradation-posterior-predictive
# | code-fold: true
# | fig-cap: Posterior predictive distribution for the transcription-splicing-degradation model.
figs = plot_sample_trajectories_with_percentiles(
    idata=idata_posterior,
)
for fig in figs:
    fig.show()
```

we can also plot the prior predictive distribution

```{python}
# | label: fig-transcription-splicing-degradation-prior-predictive
# | code-fold: true
# | fig-cap: Prior predictive distribution for the transcription-splicing-degradation model.
figs = plot_sample_trajectories_with_percentiles(
    idata=idata_posterior,
)
for fig in figs:
    fig.show()
```
